### 2025-03-03 paper.tex
keep on writing

\documentclass[fontsize=11pt]{article}
\usepackage[margin=0.75in]{geometry}
\usepackage{color}
\usepackage{graphicx}
\usepackage{float}
\usepackage{polynom}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{gensymb}
\usepackage{tikz}
\tikzset{every picture/.style={font=\normalsize}}
\DeclareOldFontCommand{\rm}{\normalfont\rmfamily}{\mathrm}
\DeclareOldFontCommand{\it}{\normalfont\rmfamily}{\mathrm}
\usepackage{mathtools}
\usepackage{hyperref}
\hypersetup{colorlinks=false,
            pdfborder={0 0 0}}
\renewcommand{\thefootnote}{\alph{footnote}}
\newcommand{\R}{\mathbb{R}}
\usepackage{amsmath, bm}   % Extra math commands and environments from the AMS
\usepackage{amssymb}   % Special symbols from the AMS
\usepackage{amsthm}    % Enhanced theorem and proof environments from the AMS
\usepackage{latexsym}  % A few extra LaTeX symbols
\usepackage{rxn}
\usepackage[version=4]{mhchem}
\usepackage{cleveref}
\usepackage{url}
\providecommand{\email}{}
\renewcommand{\email}[1]{\texttt{#1}}
\usepackage{caption}
\captionsetup{labelfont=bf, labelsep=period, labelformat=simple, labelsep=colon, figurename=Fig.}
\usepackage[authoryear, round]{natbib}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{property}[theorem]{Property}

%\newcommand{\tL}{\tilde{L}}

\graphicspath{{./}{./figures/}{./figures/paper/}}

\setcounter{topnumber}{2}              %% 2
\setcounter{bottomnumber}{1}           %% 1
\setcounter{totalnumber}{3}            %% 3
\renewcommand{\topfraction}{0.9}       %% 0.7
\renewcommand{\bottomfraction}{0.9}    %% 0.3
\renewcommand{\textfraction}{0.1}      %% 0.2
\renewcommand{\floatpagefraction}{.7}  %% 0.5

\title{\bf{Neural Ordinary Differential Equations}}
\author{Prithvi Dake\\
Department of Chemical Engineering, University of California, Santa Barbara}
\date{\textit{ME255NN: Neural Networks for modeling, optimization and control} \\
\textit{Winter 2025}}

\begin{document}

\maketitle
\section{Introduction}

Conventional neural network architectures like RNNs and residual network
resemble explicit Euler scheme to solve ODEs. Thus NODEs can be viewed as 
the \textit{continuous} versions of such \textit{discretized} architectures.
Now one problem with the continuous version is propagating the gradients during
back-propagation while training. One way to solve this problem is augmenting
the ODE with its sensitivity equations. Such a scheme saves precious memory that
goes in storing intermediate gradients in brute-force back-prop. Thus, after
overcoming the training problem, we can extend the NODE architecture to other 
applications like the continuous version of normalizing flow and state-space models.

Literature review (6 articles atleast)

The report is structured as follows: We explain conventional NN architectures like
RNN and ResNet and connect it to its continuous version called as NODE. We also show
how we can handle back-prop for NODE. We also discuss continuous normalizing flows and
NODE-based state-space models followed by some interesting case-studies. We finally 
end the report with some critical conclusions and review of current developments.

\section{NODEs and its application}
\section{Case studies}
\section{Conclusions}

\newpage
\FloatBarrier
\bibliographystyle{abbrvnat}
\bibliography{abbreviations,articles,books,unpub,proceedings}
\end{document}

### 2025-03-03 ABC_deter.py
ok we are cooking now

@@ -1,0 +1,120 @@
+import torch
+import torch.nn as nn
+import torch.optim as optim
+import torchdiffeq
+import numpy as np
+import matplotlib.pyplot as plt
+import time
+
+class ABC(nn.Module):
+    def __init__(self, config=None):
+        super(ABC, self).__init__()
+        if config is None:
+            config = {}
+        self.k1 = 1
+        self.k_1 = 1/3
+        self.k2 = 0.1
+        self.ca0 = 1
+        self.cb0 = 0
+        self.cc0 = 0
+        self.t_start = 0
+        self.t_end = 6
+        self.nplot = 75
+        self.dtype = torch.float32
+        self.mean = 0
+        self.std = 0.05
+        self.std_c = torch.tensor([0.05] * 3)
+        self.plant_solve = True
+        self.nn =  nn.Sequential(
+            nn.Linear(3, 8),
+            nn.Tanh(),
+            nn.Linear(8, 2)
+        )
+        self.nn.apply(self.init_weights)
+
+     
+    @staticmethod
+    def init_weights(m):
+        if isinstance(m, nn.Linear):
+            torch.nn.init.uniform_(m.weight, -1e-3, 1e-3).float()
+
+    def rate_true(self, y):
+        ca, cb, cc = y
+        r1 = self.k1 * ca - self.k_1 * cb
+        r2 = self.k2 * cb
+        return r1, r2
+    
+    def rate_nn(self, y):
+        r = self.nn(y)
+        return r
+
+    def plant(self, t, y):
+        ca, cb, cc = y
+        if self.plant_solve:
+            r1, r2 = self.rate_true(y)
+        else:
+            r = self.rate_nn(y)
+            r1 = r[0]
+            r2 = r[1]
+        dca_dt = -r1
+        dcb_dt = 2 * r1 - r2
+        dcc_dt = r2
+        return torch.stack([dca_dt, dcb_dt, dcc_dt])
+    
+    def solve_plant(self):
+        self.y0 = torch.tensor([self.ca0, self.cb0, self.cc0], dtype = self.dtype)
+        t = torch.linspace(self.t_start, self.t_end, self.nplot, dtype = self.dtype)
+        y = torchdiffeq.odeint(self.plant, self.y0, t, method = 'dopri5')
+        return t, y
+
+    def measurement(self):
+        t, y = self.solve_plant()
[…]

### 2025-03-05 ABC_deter.py
got pdf working

@@ -5,6 +5,8 @@ import torchdiffeq
 import numpy as np
 import matplotlib.pyplot as plt
 import time
+import scipy.stats as stats
+from mpl_toolkits.mplot3d import Axes3D
 
 class ABC(nn.Module):
     def __init__(self, config=None):
@@ -14,15 +16,15 @@ class ABC(nn.Module):
         self.k1 = 1
         self.k_1 = 1/3
         self.k2 = 0.1
-        self.ca0 = 1
-        self.cb0 = 0
-        self.cc0 = 0
+        self.ca0 = 4
+        self.cb0 = 1
+        self.cc0 = 1
         self.t_start = 0
         self.t_end = 6
         self.nplot = 75
         self.dtype = torch.float32
         self.mean = 0
-        self.std = 0.05
+        self.std = 0.1
         self.std_c = torch.tensor([0.05] * 3)
         self.plant_solve = True
         self.nn =  nn.Sequential(
@@ -101,7 +103,55 @@ class ABC(nn.Module):
 
     def multi_solve(self, ntimes = 100):
         y0 = torch.normal(mean = self.y0.expand(ntimes, -1), std = self.std_c.expand(ntimes, -1))
-        self.y0_multi = y0
+        self.plant_solve = True
+        self.y0_multi = y0.detach()
+        self.ca0_multi = np.zeros((ntimes, self.nplot))
+        self.cb0_multi = np.zeros((ntimes, self.nplot))
+        self.cc0_multi = np.zeros((ntimes, self.nplot))
+        self.y0_pdf = torch.tensor(stats.multivariate_normal.pdf(self.y0_multi, mean = self.y0, cov = self.std_c * np.eye(self.y0.shape[0])), dtype = self.dtype)
+        for i in range(ntimes):
+            self.y0 = self.y0_multi[i, :]
+            _, y = self.measurement()
+            self.ca0_multi[i, :] = y[:, 0]
+            self.cb0_multi[i, :] = y[:, 1]
+            self.cc0_multi[i, :] = y[:, 2]
+
+        axis=0
+        mean_y0 = self.y0_multi.mean(axis=axis)
+        std_y0 = self.y0_multi.std(axis=axis)
+        self.y0_norm_multi = (self.y0_multi - mean_y0) / std_y0
[…]

### 2025-03-17 ABC_plot.py
add everything to CNF repo

@@ -1,0 +1,50 @@
+# [depends] %LIB%/ABC_plant.py
+# [depends] ABC_solve.pickle
+
+import matplotlib.pyplot as plt
+import numpy as np
+from matplotlib.backends.backend_pdf import PdfPages
+import pickle
+
+figsize = (4,3)
+data = pickle.load(open('ABC_solve.pickle', 'rb'))
+t = data['t']
+y_meas = data['y_meas']
+y_pred = data['y_pred']
+y = data['y']
+y[:,:,3] = np.exp(y[:,:,3])
+
+with PdfPages('ABC_plot.pdf') as pdf:
+
+    plt.figure(figsize=figsize)
+    plt.plot(t, y_meas[:,0], 'o', color='red', mfc='none')
+    plt.plot(t, y_pred[:,0], '-', color='red')
+    plt.plot(t, y_meas[:,1], 'o', color='blue', mfc='none')
+    plt.plot(t, y_pred[:,1], '-', color='blue')
+    plt.plot(t, y_meas[:,2], 'o', color='green', mfc='none')
+    plt.plot(t, y_pred[:,2], '-', color='green')
+    plt.xlabel('$t$')
+    plt.ylabel('$C$', rotation=0)
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()
+
+    T, _ = np.meshgrid(t, np.arange(y.shape[1]), indexing='ij')  # (n_timesteps, n_samples)
+    fig, axes = plt.subplots(1, 3, figsize=(15, 5), subplot_kw={"projection": "3d"})  # 3D subplots
+    label = {'0': '$c_A$', '1': '$c_B$', '2': '$c_C$'}
+    for i, ax in enumerate(axes):
+        Y = y[:, :, i]  # Y-values (n_timesteps, n_samples)
+        Z = y[:, :, 3]  # Z-values (n_timesteps, n_samples)
+        
+        # Wireframe plot (mesh without color fill)
+        ax.plot_wireframe(T, Y, Z, color='blue', linewidth=0.5)
+        #ax.contourf(T, Y, Z, zdir='x', offset=0, cmap='coolwarm', alpha=0.5)
+        ax.view_init(azim=-20)
+        # Label axes
+        ax.set_xlabel('$t$')
+        ax.set_ylabel(f'{label[str(i)]}')
+        ax.set_zlabel('$p(c, t)$')
+    plt.tight_layout(pad = 3)
+    pdf.savefig()
+    plt.show()
+    plt.close()

### 2025-03-17 ABC_solve.py

@@ -1,0 +1,30 @@
+# [depends] %LIB%/ABC_plant.py
+
+import sys
+sys.path.append('lib/')
+from ABC_plant import ABC 
+import pickle
+
+abc = ABC()
+# get measurements
+t, y_meas = abc.measurement()
+# train the model
+abc.train(niter=500)
+# get predictions
+_, y_pred = abc.lsim(use_measure_y0=True)
+
+# generate data for CNF
+abc.multi_solve(ntimes=10)
+# train the CNF
+abc.cnf_train(niter=500)
+
+data = {}
+data['t'] = t
+data['y_meas'] = y_meas
+data['y_pred'] = y_pred
+data['y'] = abc.y
+
+with open('ABC_solve.pickle', 'wb') as f:
+    pickle.dump(data, f)
+
+

### 2025-03-17 lib/ABC_plant.py

@@ -1,0 +1,255 @@
+# Author: Prithvi Dake
+# Course: ME255NN
+
+import torch
+import torch.nn as nn
+import torch.optim as optim
+import torchdiffeq
+import numpy as np
+import matplotlib.pyplot as plt
+import time
+import scipy.stats as stats
+from mpl_toolkits.mplot3d import Axes3D
+
+class ABC(nn.Module):
+    def __init__(self, config=None):
+        '''
+        We study following kinetics: A <-> B -> C
+        with rate constants k1, k_1, k2. The code fits a neural ODE and 
+        also attempts to perform forward supervised CNF (which is not quite the
+        same as the original CNF proposed by Chen et al. 2018).
+        Link: https://proceedings.neurips.cc/paper_files/paper/2018/file/69386f6bb1dfed68692a24c8686939b9-Paper.pdf
+
+        Maybe I can call my model a generative model.
+        '''
+        super(ABC, self).__init__()
+        if config is None:
+            config = {}
+        # Default parameters
+        self.k1 = 1
+        self.k_1 = 1/3
+        self.k2 = 0.1
+        self.ca0 = 4
+        self.cb0 = 1
+        self.cc0 = 1
+
+        # Trajectory start, end and number of points
+        self.t_start = 0
+        self.t_end = 6
+        self.nplot = 75
+
+        # Data type
+        self.dtype = torch.float32
+        self.method = 'euler'
+        # Can try: 'euler', 'rk4', 'dopri5'
+
+        # Noise parameters
+        self.mean = 0
+        self.std = 0.1
+        self.std_c = torch.tensor([self.std * 2.5] * 3)
+        self.plant_solve = True
+
+        # Neural ODE
+        self.nn =  nn.Sequential(
+            nn.Linear(3, 8),
+            nn.Tanh(),
+            nn.Linear(8, 2)
+        )
+        self.nn.apply(self.init_weights)
+
+        # CNF
+        self.cnf =  nn.Sequential(
+            nn.Linear(3 , 8),
+            nn.Tanh(),
+            nn.Linear(8, 3)
+        )
+        self.cnf.apply(self.init_weights)
+
+     
+    @staticmethod
+    def init_weights(m):
+        '''
+        Initialize weights of the neural network
+        '''
+        if isinstance(m, nn.Linear):
+            torch.nn.init.uniform_(m.weight, -1e-3, 1e-3).float()
+
+    def rate_true(self, y):
+        '''
+        True rate equations
+        '''
+        ca, cb, cc = y
[…]

### 2025-03-17 ABC_deter.py
removed redundant files

@@ -1,224 +1,0 @@
-import torch
-import torch.nn as nn
-import torch.optim as optim
-import torchdiffeq
-import numpy as np
-import matplotlib.pyplot as plt
-import time
-import scipy.stats as stats
-from mpl_toolkits.mplot3d import Axes3D
-
-class ABC(nn.Module):
-    def __init__(self, config=None):
-        super(ABC, self).__init__()
-        if config is None:
-            config = {}
-        self.k1 = 1
-        self.k_1 = 1/3
-        self.k2 = 0.1
-        self.ca0 = 4
-        self.cb0 = 1
-        self.cc0 = 1
-        self.t_start = 0
-        self.t_end = 6
-        self.nplot = 75
-        self.dtype = torch.float32
-        self.mean = 0
-        self.std = 0.1
-        self.std_c = torch.tensor([0.05] * 3)
-        self.plant_solve = True
-        self.nn =  nn.Sequential(
-            nn.Linear(3, 8),
-            nn.Tanh(),
-            nn.Linear(8, 2)
-        )
-        self.nn.apply(self.init_weights)
-
-     
-    @staticmethod
-    def init_weights(m):
-        if isinstance(m, nn.Linear):
-            torch.nn.init.uniform_(m.weight, -1e-3, 1e-3).float()
-
-    def rate_true(self, y):
-        ca, cb, cc = y
-        r1 = self.k1 * ca - self.k_1 * cb
-        r2 = self.k2 * cb
-        return r1, r2
-    
-    def rate_nn(self, y):
-        r = self.nn(y)
-        return r
-
-    def plant(self, t, y):
-        ca, cb, cc = y
-        if self.plant_solve:
-            r1, r2 = self.rate_true(y)
-        else:
-            r = self.rate_nn(y)
-            r1 = r[0]
-            r2 = r[1]
-        dca_dt = -r1
-        dcb_dt = 2 * r1 - r2
-        dcc_dt = r2
-        return torch.stack([dca_dt, dcb_dt, dcc_dt])
-    
-    def solve_plant(self):
-        self.y0 = torch.tensor([self.ca0, self.cb0, self.cc0], dtype = self.dtype)
-        t = torch.linspace(self.t_start, self.t_end, self.nplot, dtype = self.dtype)
[…]

### 2025-03-19 paper.tex
more adds keep writing

@@
 \usepackage[version=4]{mhchem}
+\usepackage{algorithm, algpseudocode}
 \usepackage{cleveref}
 \usepackage{url}
+\newcommand{\norm}[1]{\lVert #1 \rVert}
 \providecommand{\email}{}
@@
 
-Conventional neural network architectures like RNNs and residual network
-resemble explicit Euler scheme to solve ODEs. Thus NODEs can be viewed as 
-the \textit{continuous} versions of such \textit{discretized} architectures.
-Now one problem with the continuous version is propagating the gradients during
-back-propagation while training. One way to solve this problem is augmenting
-the ODE with its sensitivity equations. Such a scheme saves precious memory that
-goes in storing intermediate gradients in brute-force back-prop. Thus, after
-overcoming the training problem, we can extend the NODE architecture to other 
-applications like the continuous version of normalizing flow and state-space models.
+Conventional neural network architectures, such as recurrent neural networks 
+(RNNs) and residual networks (ResNets), can be viewed as discrete approximations
+of ordinary differential equations (ODEs), specifically resembling the explicit
+Euler scheme. Neural ODEs (NODEs) extend these architectures into a continuous 
+framework, providing a more principled way of modeling dynamical systems.
 
-Literature review (6 articles atleast)
+One major challenge in training NODEs is gradient propagation during backpropagation.
+Unlike traditional architectures, where intermediate states are stored for
+gradient computation, NODEs require solving an augmented ODE system, including 
+sensitivity equations, to compute gradients efficiently. This adjoint sensitivity 
+method significantly reduces memory requirements compared to brute-force 
+backpropagation. Overcoming this training limitation enables NODEs to be 
+applied in broader contexts, such as continuous normalizing flows and state-space 
+models.
 
-The report is structured as follows: We explain conventional NN architectures like
-RNN and ResNet and connect it to its continuous version called as NODE. We also show
-how we can handle back-prop for NODE. We also discuss continuous normalizing flows and
-NODE-based state-space models followed by some interesting case-studies. We finally 
-end the report with some critical conclusions and review of current developments.
+In this review, we cover fundamental concepts and recent advancements in NODEs. 
+We begin with a discussion on artificial neural networks (ANNs) implemented in 
+a physics-informed manner that does not involve recurrence or differential 
+modeling as shown by \cite{raissi:perdikaris:karniadakis:2019}. Next, we explore black-box 
+and discrete data-driven models for process systems using ResNets and RNNs. 
+Building on these foundations, we introduce sensitivity and adjoint equation 
+methods referring \cite{stapor:froehlich:hasenauer:2018}, and then introduce the 
+formal development of NODEs by \cite{chen:rubanova:bettencourt:duvenaud:2018}.
 
-\section{NODEs and its application}
-\section{Case studies}
+We then review advanced applications, including generative latent models by
+\cite{rubanova:chen:duvenaud:2019}, continuous normalizing flows by
+\cite{grathwohl:chen:bettencourt:sutskever:duvenaud:2018}, and robustness 
+analyses of NODEs by \cite{yan:du:tan:feng:2019}. Finally, we discuss efficient 
+NODE training techniques compiled in \cite{finlay:jacobsen:nurbekyan:oberman:2020}.
+
+The structure of this report is as follows: We first introduce conventional 
+neural network architectures, such as RNNs and ResNets, and their connection 
+to NODEs. Next, we explain how backpropagation is handled in NODEs and discuss 
+applications in continuous normalizing flows and NODE-based state-space models. 
+This is followed by case studies demonstrating practical implementations. 
+The report concludes with a summary of key findings and an overview of ongoing 
+developments in the field.
+
+\section{Background}
+
+\subsection{ANNs for Physics-Informed Learning}
+
+In most engineering systems, we only have access to noisy measurements of 
+input-output data, typically originating from high-dimensional, nonlinear dynamic 
+systems. While such systems can be approximated using finite impulse response (FIR) 
+or autoregressive exogenous (ARX) models, these approaches often fail to respect 
+fundamental physical laws. One way to address this limitation is by penalizing 
+violations of these laws—a key idea behind physics-informed neural networks (PINNs), 
+as introduced by \cite{raissi:perdikaris:karniadakis:2019}.
+
+However, it is important to view PINNs not merely as data-driven models but 
+rather as a solution strategy for PDEs and ODEs. Fundamentally, they operate 
+as \textit{methods of weighted residuals}—specifically, collocation schemes—where 
+the residual is minimized at selected collocation points. While PINNs can be 
+used to learn certain model parameters, their results can be difficult to interpret 
+without a fundamental understanding of the underlying physics.
+
+Recent research has leveraged the PINN framework to enforce simple mass and energy 
+balance equations at steady-state, as this is often the only available information 
+in practical scenarios. \Cref{alg:PINNs} outlines the basic algorithm for training 
+a PINN applied to an ODE system.
+
+\begin{algorithm}[h]
+\caption{Training a Physics-Informed Neural Networks (PINNs)}
+\label{alg:PINNs}
+\begin{algorithmic}[1]
+\State \textbf{Input:} Neural network architecture $f_{\theta}$
+\State \textbf{Physical Model:} $Lu=f \quad B_1 u = 0 \quad B_2 u = 0$
+\State \textbf{Initialize} neural network parameters $\theta$
+\Repeat
+    \State $\hat{u} = f_{\theta}(x,t)$
+    \State $V_{ODE} = \norm{L\hat{u} - f}_{2}$ 
+    \Comment{Prediction error minimization} 
+    \State $V_{BC} = \norm{B_1 \hat{u}}_{2} + \norm{B_2 \hat{u}}_{2}$
+    \Comment{Enforce boundary conditions}
+    \State $V = \lambda_1 V_{PDE}  + \lambda_2 V_{BC}$
+    \Comment{Combine loss terms}
+    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} \mathcal{L}$
+    \Comment{Update $\theta$ using gradient descent}
+\Until{convergence criteria is met}
+\State \textbf{Return} trained model $f_{\theta}$
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{ResNets and RNNs for Process Systems}
+
+Most engineering systems operating near a steady state can be effectively modeled 
+using discrete linear time-invariant (DLTI) systems. These state-space models are 
+particularly well-suited for systems experiencing small perturbations.
+For highly nonlinear systems, ResNets and RNNs offer a similar state-space modeling 
+approach. In the system identification community, ResNet and RNN models are commonly 
+employed for this purpose, as illustrated in \cref{eq:resrnn}.
+
+\begin{align}
+    \hat{x}^+ &= \hat{x} + f_{\theta_{NN}} (\hat{x}, u) &\qquad \hat{x}^+ = f_{\theta_{NN}} (\hat{x}, u)\notag \\ 
+    \hat{y} &= g_{\theta_{NN}} (\hat{x}, u) &\qquad \hat{y} = g_{\theta_{NN}} (\hat{x}, u)
+    \label{eq:resrnn}
+\end{align}
+
+There are two approaches to learning the parameters $\theta_{NN}$ in \cref{eq:resrnn}: 
+\textit{multi-step prediction error minimization} and \textit{single-step prediction 
+error minimization}. 
+The single-step approach minimizes the prediction error at each time step independently, 
+while the multi-step approach considers the accumulated error over multiple time steps. 
+In practice, the multi-step method tends to be more stable, especially in the presence 
+of noisy measurements.
+
+Consider following \textit{true} model of linear chemical kinetics taking place
+in a well-mixed batch reactor with $c = (c_A, c_B, c_C)^T$ and initial concentration
+$c_0$:
+\begin{rxn*}{} 
+A \rlh[k_1][k_{-1}] 2 B, \qquad  B \rarrow[k_2] C
+\label{rxn:atobtoc}
+\end{rxn*}
+\begin{gather}
+\begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \end{bmatrix}
+=\begin{bmatrix}
+    -k_{1} & k_{-1} & 0 \\
+    2k_{1} & 2k_{-1} - k_2 & 0 \\
+    0 & k_{2} & 0
+\end{bmatrix}
+\begin{bmatrix}  c_A \\ c_B \\ c_C \end{bmatrix} 
+\label{eq:atobtoc}
+\end{gather}
+Assume we have full-state measurement. Thus, a ResNet model can be easily trained
+to predict the concentration profile over time as shown in \cref{fig:resnet}. A similar
+example can be reproduced using an RNN model. Note we used multistep prediction error
+minimization. One disadvantage of such iterative models is that they require large memory
+to store intermediate states for gradient computation during backpropagation (since we basically
+apply chain-rule).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.55\textwidth, page=2]{ABC_plot.pdf} 
+    \caption{Concentration profiles of species A, B, and C in a batch reactor. The 
+    discrete ResNet model does quite a good job to predict the concentration profiles.}
+    \label{fig:resnet}
+\end{figure}
+
+\subsection{Sensitivity or Adjoint Equation Methods}
+\label{sec:sens}
+
+Here we give a quick overview of the sensitivity equations for ODEs. Consider following
+non-linear state-space model:
+
+\begin{align}
+    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag \\
+    y &= g(x, u; \theta)
+\end{align}
+
+Assume $y=x$ i.e. full state measurement and hence we also know the initial condition
+$x_0$. Consider a least-squares objective to measure the fit to the data:
+
+\begin{align}
+    \min\limits_{\theta} V(x, u; \theta) = \norm{y - \hat{y}}^2 
+\end{align}
+To propagate the gradient we need to solve the sensitivity equations:
+\begin{align}
+    \dfrac{\partial V(x, u; \theta)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
+    S_{ij} &= \dfrac{\partial x_i}{\partial \theta_j}
+    \label{eq:sens}
+\end{align}
+where $S_{ij}$ is the sensitivity of model solution with parameter. We differentiate $S_{ij}$ to obtain:
+\begin{align}
+    \dfrac{d S_{ij}}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x_i}{\partial \theta_j}\right) =
+    \dfrac{\partial}{\partial\theta_j} \left(\dfrac{d x_i}{dt}\right) = 
+    \dfrac{\partial }{\partial \theta_j} f_i(x; \theta) 
+\end{align}
+Using chain rule on $f$, we get following matrix equation:
+\begin{align}
+    \dfrac{dS}{dt} = f_xS + f_{\theta} \qquad  S(0) = g_\theta
+\end{align}
+We could easily calculate the initial condition 
+if we are using a state estimator by using the \cref{eq:sens}. Thus, we can construct
+following augmented ODE system:
+\begin{align}
+    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
+    \begin{bmatrix} f(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
+    \qquad
+    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
+    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
+    \label{eq:augsens}
+\end{align}
+This augmented system can be solved using any ODE solver, allowing gradient 
+propagation without the need to store intermediate states. This key advantage 
+contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
+as a continuous counterpart to Euler-step-based architectures like ResNets. We close 
+this section by a short example on solving the augmented ODE system. Consider, 
+following kinetics:
+\begin{align}
+    \dot{c}_A &= -kc_A \notag \\
+    c_A(0) &= c_{A0}
+\end{align}
+Say we want to calculate sensitivities wrt $k, c_{A0}$. With reference to \cref{eq:augsens},
+for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
+we have following augmented system:
+\begin{align}
+    \begin{bmatrix} 
+        \dot{c}_A \\
+        \dot{S_1} \\ 
+        \dot{S_2} 
+    \end{bmatrix} = 
+    \begin{bmatrix}
+        -kc_A \\
+        -kS_1 - c_A \\
+        -kS_2 \end{bmatrix}
+    \qquad
+    \begin{bmatrix} 
+        c_A(0) \\ 
+        S_1(0) \\ 
+        S_2(0) 
+    \end{bmatrix} =
+    \begin{bmatrix} 
+        c_{A0} \\ 
+        0 \\ 
+        1
+    \end{bmatrix}
+\end{align}
+If we were learning the parameter $k$ and initial condition $c_{A0}$, we could
+show a gradient descent update as follows:
+\begin{align}
+    \begin{bmatrix} k \\ c_{A0} \end{bmatrix} \leftarrow
+    \begin{bmatrix} k \\ c_{A0} \end{bmatrix}
+    - \eta 
+    \begin{bmatrix} 
+        -2 (\hat{c}_A  - c_A)^T S_1 \\ 
+        -2 (\hat{c}_A  - c_A)^T S_2
+    \end{bmatrix}
+\end{align}
+Thus, we do no need to backprop but just solve the augmented ODE system once 
+in each optimal iteration unlike ResNets or RNNs. However, such a method is not 
+an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
+numerical ODE solvers like stiffness, accuracy, etc. Sometimes we may end up 
+with noisy gradients which may lead to
+poor convergence.
+
+\subsection{Neural Ordinary Differential Equations}
+
+With the background established in \cref{sec:sens}, we can now introduce the algorithm
+for training NODEs. The key idea is to replace the right-hand side of the ODE system
+with a neural network. With auto-differentiation we can easily construct the augmented
+ODE system and solve it using any ODE solver.
+
+\begin{algorithm}[h]
+\caption{Training a Neural Ordinary Differential Equation (NODE)}
+\label{alg:NODE}
+\begin{algorithmic}[1]
+\State \textbf{Input:} Neural network architecture $f_{\theta}$
+\State \textbf{ODE:} 
+$
+    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
+    \begin{bmatrix} f_\theta(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
+    \qquad
+    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
+    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
+    \label{eq:augsens}
+$
+\Comment{Construct the augmented ODE system say, using \texttt{PyTorch}}
+\State \textbf{Initialize} neural network parameters $\theta$
+\Repeat
+    \State $\hat{x}_{aug} = \text{ODESolve}(f_{aug}, x_{aug}(0), t)$
+    \State $V = \norm{\hat{x} - x}_{2}$
+    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} \mathcal{L}$
+\Until{convergence criteria is met}
+\State \textbf{Return} trained model $f_{\theta}$
+\end{algorithmic}
+\end{algorithm}
+We close the section with a same example as shown in \cref{eq:atobtoc} but now
+we use a NODE to predict the concentration profiles. The results are shown in
+\cref{fig:node}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.55\textwidth, page=1]{ABC_plot.pdf} 
+    \caption{Concentration profiles of species A, B, and C in a batch reactor. The
+    NODE model also does quite a good job to predict the concentration profiles. 
+    In fact for such simple problem, we hardly see any difference between ResNet and NODE.}
+    \label{fig:node}
+\end{figure}
+
+\subsection{Generative latent models based on continuous normalizing flows}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
+\end{figure}
+
+\subsection{Robustness analysis of NODEs}
+
+\subsection{Efficient training techniques for NODEs}
+
 \section{Conclusions}
@@
 \bibliographystyle{abbrvnat}
-\bibliography{abbreviations,articles,books,unpub,proceedings}
+\bibliography{cnf_bib, abbreviations,articles,books,unpub,proceedings}
 \end{document}

### 2025-03-19 ABC_plot.py

@@ -11,6 +11,7 @@ data = pickle.load(open('ABC_solve.pickle', 'rb'))
 t = data['t']
 y_meas = data['y_meas']
 y_pred = data['y_pred']
+y_pred_res = data['y_pred_res']
 y = data['y']
 y[:,:,3] = np.exp(y[:,:,3])
 
@@ -29,8 +30,21 @@ with PdfPages('ABC_plot.pdf') as pdf:
     pdf.savefig()
     plt.close()
 
+    plt.figure(figsize=figsize)
+    plt.plot(t, y_meas[:,0], 'o', color='red', mfc='none')
+    plt.plot(t, y_pred_res[:,0], '-', color='red')
+    plt.plot(t, y_meas[:,1], 'o', color='blue', mfc='none')
+    plt.plot(t, y_pred_res[:,1], '-', color='blue')
+    plt.plot(t, y_meas[:,2], 'o', color='green', mfc='none')
+    plt.plot(t, y_pred_res[:,2], '-', color='green')
+    plt.xlabel('$t$')
+    plt.ylabel('$C$', rotation=0)
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()
+
     T, _ = np.meshgrid(t, np.arange(y.shape[1]), indexing='ij')  # (n_timesteps, n_samples)
-    fig, axes = plt.subplots(1, 3, figsize=(15, 5), subplot_kw={"projection": "3d"})  # 3D subplots
+    fig, axes = plt.subplots(1, 3, figsize=(9, 3), subplot_kw={"projection": "3d"})  # 3D subplots
     label = {'0': '$c_A$', '1': '$c_B$', '2': '$c_C$'}
     for i, ax in enumerate(axes):
         Y = y[:, :, i]  # Y-values (n_timesteps, n_samples)
@@ -39,12 +53,17 @@ with PdfPages('ABC_plot.pdf') as pdf:
         # Wireframe plot (mesh without color fill)
         ax.plot_wireframe(T, Y, Z, color='blue', linewidth=0.5)
         #ax.contourf(T, Y, Z, zdir='x', offset=0, cmap='coolwarm', alpha=0.5)
-        ax.view_init(azim=-20)
+        ax.view_init(azim=20)
         # Label axes
         ax.set_xlabel('$t$')
         ax.set_ylabel(f'{label[str(i)]}')
-        ax.set_zlabel('$p(c, t)$')
-    plt.tight_layout(pad = 3)
+        if i == 0:
[…]

### 2025-03-19 ABC_solve.py

@@ -13,6 +13,11 @@ abc.train(niter=500)
 # get predictions
 _, y_pred = abc.lsim(use_measure_y0=True)
 
+abc.resnet_train(niter=500)
+# get predictions
+_, y_pred_res = abc.resnet_lsim(use_measure_y0=True)
+
+
 # generate data for CNF
 abc.multi_solve(ntimes=10)
 # train the CNF
@@ -22,6 +27,7 @@ data = {}
 data['t'] = t
 data['y_meas'] = y_meas
 data['y_pred'] = y_pred
+data['y_pred_res'] = y_pred_res
 data['y'] = abc.y
 
 with open('ABC_solve.pickle', 'wb') as f:

### 2025-03-19 lib/ABC_plant.py

@@ -57,6 +57,13 @@ class ABC(nn.Module):
         )
         self.nn.apply(self.init_weights)
 
+        self.resnet = nn.Sequential(
+            nn.Linear(3, 8),
+            nn.Tanh(),
+            nn.Linear(8, 3)
+        )
+        self.resnet.apply(self.init_weights)
+
         # CNF
         self.cnf =  nn.Sequential(
             nn.Linear(3 , 8),
@@ -156,6 +163,60 @@ class ABC(nn.Module):
                 print(f'Iter {i}, Loss {loss.item()}')
         toc = time.time()
         print(f'Training time: {(toc - tic)/60} mins')
+
+
+    def resnet_plant(self, use_measure_y0=False, y0=None):
+        '''
+        Get the loss to train the resnet
+        '''
+        t = torch.linspace(self.t_start, self.t_end, self.nplot, dtype = self.dtype)
+        if use_measure_y0:
+            yt0 = self.measure[0, :]
+        elif y0 is not None:
+            yt0 = y0
+        else:
+            yt0 = torch.tensor([self.ca0, self.cb0, self.cc0], dtype = self.dtype)
+        
+        y_sol = [yt0]
+        for i in range(len(t) - 1):
+            next_y = self.resnet(y_sol[-1]) + y_sol[-1]
+            y_sol.append(next_y)
+        y_sol = torch.stack(y_sol)
+        #loss = torch.mean((y_sol - self.measure)**2)
+        return t, y_sol
+    
+    def resnet_loss(self, use_measure_y0=False):
+        '''
+        Get the loss to train the resnet
+        '''
+        t, y_sol = self.resnet_plant(use_measure_y0=use_measure_y0)
+        loss = torch.mean((y_sol - self.measure)**2)
+        return loss
+    
+    def resnet_train(self, niter = 1000):
+        '''
+        Train the resnet
+        '''
+        optimizer = optim.Adam(self.resnet.parameters(), lr=1e-2)
+        tic = time.time()
+        for i in range(niter):
+            optimizer.zero_grad()
+            loss = self.resnet_loss(use_measure_y0=True)
+            loss.backward()
+            optimizer.step()
+            if i % 100 == 0:
+                print(f'Iter {i}, Loss {loss.item()}')
+        toc = time.time()
+        print(f'Training time: {(toc - tic)/60} mins')
+
+    def resnet_lsim(self, use_measure_y0=False, y0=None):
+        '''
[…]

### 2025-03-19 paper.tex
added derivation to report

@@
 
-\subsection{Sensitivity or Adjoint Equation Methods}
+\subsection{Forward and backward Sensitivity/Adjoint methods}
+
+\subsubsection{Forward Sensitivity}
 \label{sec:sens}
 
-Here we give a quick overview of the sensitivity equations for ODEs. Consider following
+Here we give a quick overview of both the forward/backward sensitivity equations for ODEs. Consider following
 non-linear state-space model:
@@
     y &= g(x, u; \theta)
+    \label{eq:ssode}
 \end{align}
@@
 \begin{align}
-    \min\limits_{\theta} V(x, u; \theta) = \norm{y - \hat{y}}^2 
+    \min\limits_{\theta} V(x) = \norm{x - \hat{x}}^2 
 \end{align}
@@
 \begin{align}
-    \dfrac{\partial V(x, u; \theta)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
+    \dfrac{\partial V(x)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
     S_{ij} &= \dfrac{\partial x_i}{\partial \theta_j}
@@
 This augmented system can be solved using any ODE solver, allowing gradient 
-propagation without the need to store intermediate states. This key advantage 
-contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
-as a continuous counterpart to Euler-step-based architectures like ResNets. We close 
-this section by a short example on solving the augmented ODE system. Consider, 
+propagation. However, if $\theta \in \R^{p}$ and $x \in \R^{n}$, we need to 
+solve $p \times n + 1$ equations i.e. the cost increases with $p$. Thus, forward
+sensitivity is not an efficient method to propagate gradients. However, we can circumvent
+the problem of storing intermediate states using backward sensitivity method discussed next. This key advantage 
+of backward adjoint method contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
+as a continuous counterpart to Euler-step-based architectures like ResNets. We also show 
+a short example on solving the augmented ODE system for forward sensitivity. Consider, 
 following kinetics:
@@
 \end{align}
-Thus, we do no need to backprop but just solve the augmented ODE system once 
-in each optimal iteration unlike ResNets or RNNs. However, such a method is not 
+
+\subsubsection{Backward Sensitivity}
+
+In forward sensitivity since we solve the augmented ODE system, we do not need 
+to worry about time-steps taken by the ODE solver. However for backward sensitivity,
+it becomes a major problem. Here we show derivation of the adjoint equations for the
+continuous version rather than the previous discrete version (where the loss is obtained by summing).
+Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:
+
+\begin{align}
+    V(x) &= \int_0^T (y - \hat{y})^2 dt = \int_0^T g(x, u; \theta)dt \notag \\
+    \mathcal{L} &= \int_0^T g dt + \int_0^T \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
+\end{align}
+Hence,
+\begin{align}
+    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
+    \int_0^T \lambda^T \left(\dfrac{\partial f}{\partial x} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta} - \dfrac{\partial}{\partial \theta}\dfrac{dx}{dt}\right) dt \notag \\
+    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
+    \int_0^T \lambda^T \left(\dfrac{\partial f}{\partial x} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta} - \dfrac{d}{dt}\dfrac{\partial x}{\partial \theta^T}\right) dt \notag \\
+    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \lambda^T \dfrac{\partial f}{\partial \theta} dt + \int_0^T \left(\dfrac{\partial g}{\partial x^T}
+    + \lambda^T \dfrac{\partial f}{\partial x} - \lambda^T\dfrac{d}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt
+\end{align}
+The term $\int_0^T{\lambda^T \dfrac{d}{dt} \dfrac{\partial x}{\partial \theta^T}dt}$ 
+can be solved using integration by parts. Thus, we have:
+
+\begin{align}
+    \dfrac{\partial \mathcal{L}}{\partial \theta} &= \int_0^T \lambda^T \dfrac{\partial f}{\partial \theta} dt + \int_0^T \left(\dfrac{\partial g}{\partial x^T}
+    + \lambda^T \dfrac{\partial f}{\partial x} - \dfrac{d\lambda}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt - \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0
+    + \lambda^T(T)\dfrac{\partial x}{\partial \theta^T}|_T 
+    \label{eq:adjoint}
+\end{align}
+
+Now, the whole objective of backward sensitivity is to avoid calculating $\dfrac{\partial x}{\partial \theta^T}$.
+Also note that the equality constraint in the Langrange functional is always satisfied
+since we solve the ODE system. Thus we get to choose $\lambda$ such that the terms
+associated with $\dfrac{\partial x}{\partial \theta^T}$ vanish. Also, we can 
+easily calculate $\dfrac{\partial x}{\partial \theta^T}|_0$ but we would like to set
+$\dfrac{\partial x}{\partial \theta^T}|_T = 0$. Thus, we can construct following adjoint 
+equations:
+
+\begin{align}
+    \dfrac{d\lambda}{dt} &= \left(\dfrac{\partial g}{\partial x^T} + \lambda^T \dfrac{\partial f}{\partial x}\right)^T \notag \\
+    \lambda(T) &= 0
+\end{align}
+Referring to \cref{eq:adjoint}, we see $\dfrac{\partial V}{\partial \theta} = \dfrac{\partial \mathcal{L}}{\partial \theta} + \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0$.
+Thus, we do no need to backprop but just solve the augmented ODE system once forward 
+and once backward in each optimal iteration unlike ResNets or RNNs. Also, note that the number 
+of adjoint equations do not increase with parameters. However, such a method is not 
 an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by

### 2025-03-20 paper.tex
geting to completion finally

--- 2025-03-19 paper.tex
@@
 \usepackage{hyperref}
+\usepackage{pgfplots}
 \hypersetup{colorlinks=false,
@@
 Unlike traditional architectures, where intermediate states are stored for
-gradient computation, NODEs require solving an augmented ODE system, including 
-sensitivity equations, to compute gradients efficiently. This adjoint sensitivity 
+gradient computation, NODEs require solving an augmented ODE system, involving
+adjoint sensitivity equations, to compute gradients efficiently. This adjoint sensitivity 
 method significantly reduces memory requirements compared to brute-force 
-backpropagation. Overcoming this training limitation enables NODEs to be 
+backpropagation \& forward sensitivity equations. Overcoming this training limitation enables NODEs to be 
 applied in broader contexts, such as continuous normalizing flows and state-space 
@@
 a physics-informed manner that does not involve recurrence or differential 
-modeling as shown by \cite{raissi:perdikaris:karniadakis:2019}. Next, we explore black-box 
-and discrete data-driven models for process systems using ResNets and RNNs. 
+modeling as shown by \cite{raissi:perdikaris:karniadakis:2019}. Next, we explore
+discrete data-driven models for process systems using ResNets and RNNs. 
 Building on these foundations, we introduce sensitivity and adjoint equation 
@@
 applications in continuous normalizing flows and NODE-based state-space models. 
-This is followed by case studies demonstrating practical implementations. 
+Most of the sections give exemplar implementations. 
 The report concludes with a summary of key findings and an overview of ongoing 
@@
 \begin{algorithmic}[1]
-\State \textbf{Input:} Neural network architecture $f_{\theta}$
-\State \textbf{Physical Model:} $Lu=f \quad B_1 u = 0 \quad B_2 u = 0$
+\State \textbf{Input:} Neural network architecture $f_{NN}$
+\State \textbf{Physical Model:} $Lx=f \quad B_1 x = 0 \quad B_2 x = 0$
+\Comment{True or known physics}
 \State \textbf{Initialize} neural network parameters $\theta$
 \Repeat
-    \State $\hat{u} = f_{\theta}(x,t)$
-    \State $V_{ODE} = \norm{L\hat{u} - f}_{2}$ 
+    \State $x = f_{NN}(x,t; \theta)$
+    \State $V_{ODE} = \norm{Lx - f}^{2}$ 
     \Comment{Prediction error minimization} 
-    \State $V_{BC} = \norm{B_1 \hat{u}}_{2} + \norm{B_2 \hat{u}}_{2}$
+    \State $V_{BC} = \norm{B_1 x}^{2} + \norm{B_2 x}^{2}$
     \Comment{Enforce boundary conditions}
@@
     \Comment{Combine loss terms}
-    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} \mathcal{L}$
+    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} V$
     \Comment{Update $\theta$ using gradient descent}
 \Until{convergence criteria is met}
-\State \textbf{Return} trained model $f_{\theta}$
+\State \textbf{Return} trained model $f_{NN}$
 \end{algorithmic}
@@
 approach. In the system identification community, ResNet and RNN models are commonly 
-employed for this purpose, as illustrated in \cref{eq:resrnn}.
+employed for this purpose, as illustrated in \cref{eq:resrnn} on left and right respectively.
 
 \begin{align}
-    \hat{x}^+ &= \hat{x} + f_{\theta_{NN}} (\hat{x}, u) &\qquad \hat{x}^+ = f_{\theta_{NN}} (\hat{x}, u)\notag \\ 
-    \hat{y} &= g_{\theta_{NN}} (\hat{x}, u) &\qquad \hat{y} = g_{\theta_{NN}} (\hat{x}, u)
+    x^+ &= x + f_{NN} (x, u; \theta) &\qquad x^+ = f_{NN} (x, u; \theta)\notag \\ 
+    y &= g_{NN} (x, u; \theta) &\qquad y = g_{NN} (x, u; \theta)
     \label{eq:resrnn}
@@
 
-There are two approaches to learning the parameters $\theta_{NN}$ in \cref{eq:resrnn}: 
+There are two approaches to learning the parameters $\theta$ in \cref{eq:resrnn}: 
 \textit{multi-step prediction error minimization} and \textit{single-step prediction 
@@
 \end{gather}
-Assume we have full-state measurement. Thus, a ResNet model can be easily trained
-to predict the concentration profile over time as shown in \cref{fig:resnet}. A similar
-example can be reproduced using an RNN model. Note we used multistep prediction error
-minimization. One disadvantage of such iterative models is that they require large memory
-to store intermediate states for gradient computation during backpropagation (since we basically
-apply chain-rule).
+Assume we have full-state measurement i.e.we measure all the three concentrations. Thus, 
+a ResNet model can be easily trained to predict the concentration profile over 
+time as shown in \cref{fig:resnet}. A similar example can be reproduced using an 
+RNN model. Note we used multi-step prediction error minimization. One disadvantage 
+of such iterative models is that they require large memory
+to store intermediate states for gradient propagation (since we basically
+apply chain-rule while differentiating through the solution). Another disadvantage is they 
+are discrete in nature and thus, can struggle when the data is not sampled at regular intervals.
 
@@
 
-\subsection{Sensitivity or Adjoint Equation Methods}
+\subsection{Forward and backward sensitivity/adjoint methods}
 \label{sec:sens}
+\subsubsection{Forward Sensitivity}
 
-Here we give a quick overview of the sensitivity equations for ODEs. Consider following
-non-linear state-space model:
-
+Here we give a quick overview of both the forward and backward sensitivity/adjoint 
+equations for ODEs. Consider following non-linear state-space model:
 \begin{align}
-    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag \\
-    y &= g(x, u; \theta)
+    \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)\notag 
+    \label{eq:ssode}
 \end{align}
-
-Assume $y=x$ i.e. full state measurement and hence we also know the initial condition
-$x_0$. Consider a least-squares objective to measure the fit to the data:
-
+We assume full state measurement (i.e. $y=x$). Consider a least-squares objective
+to measure the fit to the data ($\tilde{x}$):
 \begin{align}
-    \min\limits_{\theta} V(x, u; \theta) = \norm{y - \hat{y}}^2 
+    \min\limits_{\theta} V(x) = \norm{\tilde{x} - x}^2 
 \end{align}
@@
 \begin{align}
-    \dfrac{\partial V(x, u; \theta)}{\partial \theta_j} &= -2 \sum_i (\hat{x} - x) \dfrac{\partial x_i}{\partial \theta_j} \notag \\ 
-    S_{ij} &= \dfrac{\partial x_i}{\partial \theta_j}
+    \dfrac{\partial V}{\partial \theta^T} &= -2 (\tilde{x} - x)^T \dfrac{\partial x}{\partial \theta^T} \notag \\ 
+    S &= \dfrac{\partial x}{\partial \theta^T}
     \label{eq:sens}
 \end{align}
-where $S_{ij}$ is the sensitivity of model solution with parameter. We differentiate $S_{ij}$ to obtain:
+where $S$ is the \textit{sensitivity} matrix of the model solutions with parameters. We differentiate $S$ to obtain:
 \begin{align}
-    \dfrac{d S_{ij}}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x_i}{\partial \theta_j}\right) =
-    \dfrac{\partial}{\partial\theta_j} \left(\dfrac{d x_i}{dt}\right) = 
-    \dfrac{\partial }{\partial \theta_j} f_i(x; \theta) 
+    \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
+    \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
+    \dfrac{\partial }{\partial \theta^T} f
 \end{align}
@@
 \begin{align}
-    \dfrac{dS}{dt} = f_xS + f_{\theta} \qquad  S(0) = g_\theta
+    \dfrac{dS}{dt} = \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \qquad  S(0) = \dfrac{\partial g}{\partial \theta^T}
 \end{align}
-We could easily calculate the initial condition 
-if we are using a state estimator by using the \cref{eq:sens}. Thus, we can construct
-following augmented ODE system:
+Thus, we can construct following augmented ODE system:
 \begin{align}
     \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
-    \begin{bmatrix} f(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
+    \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
     \qquad
     \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
-    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
-    \label{eq:augsens}
+    \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
+    \label{eq:augsensode}
 \end{align}
 This augmented system can be solved using any ODE solver, allowing gradient 
-propagation without the need to store intermediate states. This key advantage 
-contributed to the popularity of Neural ODEs (NODEs), beyond their interpretation 
-as a continuous counterpart to Euler-step-based architectures like ResNets. We close 
-this section by a short example on solving the augmented ODE system. Consider, 
+propagation. However, if $\theta \in \R^{p}$ and $x \in \R^{n}$, we need to 
+solve $S \in \R^{n+p}$ equations i.e. the cost increases with $p$. Thus, forward
+sensitivity is not an efficient method to propagate gradients. However, we can circumvent
+the problem of linear increase in sensitivity equations using backward adjoint equations. This key advantage 
+of backward adjoint method contributed to the feasibility of Neural ODEs (NODEs), beyond their interpretation 
+as a continuous counterpart to Euler-step based architectures like ResNets. We next show 
+a short example on solving the augmented ODE system for forward sensitivity. Consider, 
 following kinetics:
@@
 \end{align}
-Say we want to calculate sensitivities wrt $k, c_{A0}$. With reference to \cref{eq:augsens},
+Say we want to calculate sensitivities wrt $k$ and $c_{A0}$. With reference to \cref{eq:augsensode},
 for $S_1 = \dfrac{\partial c_A}{\partial k}$ and $S_2 = \dfrac{\partial c_A}{\partial c_{A0}}$,
@@
 \end{align}
-Thus, we do no need to backprop but just solve the augmented ODE system once 
-in each optimal iteration unlike ResNets or RNNs. However, such a method is not 
+
+\subsubsection{Backward Sensitivity}
+
+In this section we redefine the objective function as an integral over time.
+The integral formulation allows us to derive the adjoint equation.
+Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:
+
+\begin{align}
+    \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \hat{x}}^2 dt = \int_0^t g(x)dt \notag \\
+    \min\limits_{\theta, \lambda}\mathcal{L} &= \int_0^t g dt + \int_0^t \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
+\end{align}
+Hence,
+\begin{align}
+    \dfrac{\partial \mathcal{L}}{\partial \theta^T} &= \int_0^t \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
+    \int_0^t \lambda^T \left(\dfrac{\partial f}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta^T} - \dfrac{\partial}{\partial \theta^T}\dfrac{dx}{dt}\right) dt \notag \\
+    \dfrac{\partial \mathcal{L}}{\partial \theta^T} &= \int_0^t \dfrac{\partial g}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} dt + 
+    \int_0^t \lambda^T \left(\dfrac{\partial f}{\partial x^T} \dfrac{\partial x}{\partial \theta^T} + \dfrac{\partial f}{\partial \theta^T} - \dfrac{d}{dt}\dfrac{\partial x}{\partial \theta^T}\right) dt \notag \\
+    \dfrac{\partial \mathcal{L}}{\partial \theta^T} &= \int_0^t \lambda^T \dfrac{\partial f}{\partial \theta^T} dt + \int_0^t \left(\dfrac{\partial g}{\partial x^T}
+    + \lambda^T \dfrac{\partial f}{\partial x^T} - \lambda^T\dfrac{d}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt
+\end{align}
+The term $\int_0^t{\lambda^T \dfrac{d}{dt} \dfrac{\partial x}{\partial \theta^T}dt}$ 
+can be solved using integration by parts. Thus, we have:
+
+\begin{align}
+    \dfrac{\partial \mathcal{L}}{\partial \theta^T} &= \int_0^t \lambda^T \dfrac{\partial f}{\partial \theta^T} dt + \int_0^t \left(\dfrac{\partial g}{\partial x^T}
+    + \lambda^T \dfrac{\partial f}{\partial x^T} + \dfrac{d\lambda^T}{dt}\right)\dfrac{\partial x}{\partial \theta^T} dt + \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0
+    - \lambda^T(T)\dfrac{\partial x}{\partial \theta^T}|_t
+    \label{eq:adjoint}
+\end{align}
+
+Now, the whole objective of backward sensitivity is to avoid calculating $\dfrac{\partial x}{\partial \theta^T}$.
+Also note that the equality constraint in the Langrange functional is always satisfied
+since we \textit{solve} the ODE system. Thus we get to choose $\lambda$ such that the terms
+associated with $\dfrac{\partial x}{\partial \theta^T}$ vanish. Also, we can 
+easily calculate $\dfrac{\partial x}{\partial \theta^T}|_0$ but we would like to set
+$\dfrac{\partial x}{\partial \theta^T}|_t = 0$. Thus, we can construct following adjoint 
+equations:
+
+\begin{align}
+    \dfrac{d\lambda^T}{dt} &= -\dfrac{\partial g}{\partial x^T} - \lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
+    \lambda(T) &= 0
+\end{align}
+Referring to \cref{eq:adjoint}, we see $\dfrac{\partial V}{\partial \theta^T} = 
+\dfrac{\partial \mathcal{L}}{\partial \theta^T}$.
+The original NODE paper uses objective function of the form:
+\begin{align}
+    \min\limits_{\theta}V(x) = \int_0^t \norm{x - \hat{x}}^2 \delta(t) dt
+\end{align}
+Hence, the adjoint equations for their case are:
+\begin{align}
+    \dfrac{d\lambda^T}{dt} &= -\lambda^T \dfrac{\partial f}{\partial x^T} \notag \\
+    \lambda(t)^T &= - \dfrac{\partial g}{\partial x^T}|_t = -\dfrac{\partial V}{\partial x^T}|_t \notag \\
+    \dfrac{\partial V}{\partial \theta^T} &= \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
+    \label{eq:adjointnode}
+\end{align}
+Hence, $\lambda = \dfrac{\partial V}{\partial x}$ as we see from the terminal 
+condition in \cref{eq:adjointnode}.
+Thus, we need to solve the augmented ODE system once forward 
+and once backward in each optimal iteration. We need a forward pass to store
+the value of $\dfrac{\partial g}{\partial x^T}|_t$ and $x_t$. Also, note that the number 
+of adjoint equations do not increase with parameters (i.e $\lambda \in \R^n$ unlike $S \in \R^{n \times p+1}$). 
+We can evaluate the gradient in one go as shown in \cref{eq:adjointnode}. 
+Some recent solutions to storing intermediate points involve \textit{checkpointing}
+at only few points. However, such a method is not 
 an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
-numerical ODE solvers like stiffness, accuracy, etc. Sometimes we may end up 
-with noisy gradients which may lead to
-poor convergence.
+numerical ODE solvers like stiffness, accuracy, etc. One can analyze the stiffness
+of the adjoint ODE by simply inspecting the eignevalues of the Jacobian matrix $\dfrac{\partial f}{\partial x^T}$
+at different time points. Sometimes we may end up with noisy gradients which may lead to poor convergence.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.55\textwidth, page=4]{ABC_plot.pdf} 
+    \caption{Consider $\dot{\lambda} = -e \lambda$ for $e = \pm 10$. 
+    We see that the adjoint variable will either explode or vanish.
+    Thus $\dfrac{\partial V}{\partial \theta} = \lambda(0) = 10^{\pm 5}$. 
+    A simple example to show how an ill-conditioned
+    adjoint system can lead to poor convergence.}
+    \label{fig:node}
+\end{figure}
 
@@
 \begin{algorithmic}[1]
-\State \textbf{Input:} Neural network architecture $f_{\theta}$
+\State \textbf{Input:} Neural network architecture $f_{NN}$
 \State \textbf{ODE:} 
 $
-    \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
-    \begin{bmatrix} f_\theta(x, u; \theta) \\ f_xS + f_{\theta} \end{bmatrix}
+    \begin{bmatrix} \dot{x} \\ \dot{\lambda^T} \end{bmatrix} = 
+    \begin{bmatrix} f_{NN} \\ -\lambda^T \dfrac{\partial f_{NN}}{\partial x^T}\end{bmatrix}
     \qquad
-    \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
-    \begin{bmatrix} x_0 \\ g_\theta \end{bmatrix}
+    \begin{bmatrix} x(t) \\ \lambda(t) \end{bmatrix} =
+    \begin{bmatrix} x_t \\ - \dfrac{\partial g}{\partial x^T}|_{t} \end{bmatrix}
     \label{eq:augsens}
 $
-\Comment{Construct the augmented ODE system say, using \texttt{PyTorch}}
+\Comment{Construct the augmented ODE using \texttt{PyTorch}}
 \State \textbf{Initialize} neural network parameters $\theta$
 \Repeat
-    \State $\hat{x}_{aug} = \text{ODESolve}(f_{aug}, x_{aug}(0), t)$
-    \State $V = \norm{\hat{x} - x}_{2}$
-    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} \mathcal{L}$
+    \State $V = \norm{\hat{x} - x}^2$
+    \Comment{Forward solve to get the loss $V$}
+    \State $\hat{x}_{aug} = \text{ODESolve}(f_{aug}, x_{aug}(0), T, t_0)$
+    \State $\dfrac{\partial V}{\partial \theta^T} = \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt$
+    \Comment{Construct the gradient wrt loss}
+    \State $\theta \leftarrow \theta - \eta \nabla_{\theta} V$
 \Until{convergence criteria is met}
-\State \textbf{Return} trained model $f_{\theta}$
+\State \textbf{Return} trained model $f_{NN}$
 \end{algorithmic}
 \end{algorithm}
-We close the section with a same example as shown in \cref{eq:atobtoc} but now
+We close the section with the same example as shown in \cref{eq:atobtoc} but now
 we use a NODE to predict the concentration profiles. The results are shown in
@@
 
+The NODE paper applies the same motivation to develop continuous normalizing flows.
+Interested reader can refer the paper for theorem on instantaneous change of variables.
+The key idea is the neural ODE rhs is trained to conserve the probability density 
+while maximizing log-likelihood to describe the data. The framework can be described 
+as follows:
+\begin{align}
+    \begin{bmatrix}
+        x_0 \\
+        \log p(x) - \log p(x_0)
+    \end{bmatrix}
+    =
+    \int_{t}^{0}
+    \begin{bmatrix}
+        f_{NN}(x, u; \theta) \\
+        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
+    \end{bmatrix} dt;
+    \qquad
+    \begin{bmatrix}
+        x(t) \\
+        \log p(x) - \log p(x_t)
+    \end{bmatrix}   
+    =
+    \begin{bmatrix}
+        x_t \\
+        0
+    \end{bmatrix}
+\end{align}
+We assume, $x_0 \sim \mathcal{N}(0, \sigma^2I)$ and thus the objective function is:
+\begin{align}
+    \min\limits_{\theta} V(x) = -\log p(x)
+\end{align}
+That's a powerful framework to model distributions (say a human face composed of pixels).
+However, we are more interested in using similar generative framework for chemical kinetics. 
+Thus, we again use the same example as shown in \cref{eq:atobtoc} but we also model 
+how the uncertainty in the initial conditions decreases as we add more measurement points.
+The results are shown in \cref{fig:cnf}. The system we solve is:
+
+\begin{align}
+    \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
+    = \begin{bmatrix}
+        -k_{1}c_A + k_{-1}c_B  \\
+        2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
+        k_{2}c_B\\
+        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
+    \end{bmatrix};
+    &\qquad
+    \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
+    \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
+    \min\limits_{\theta} V(x) &= -\log p(x) + \norm{c - \tilde{c}}^2
+\end{align}
+The author would like to draw parallels to the classical Kalman filter that also does the similar job,
+but the continuous normalizing flow can be applied to any non-linear dynamical system.
 \begin{figure}[h]
@@
     \includegraphics[width=1\textwidth, page=3]{ABC_plot.pdf} 
+    \caption{Results for chemical kinetics modeled as continuous normalizing
+    flow. We start with a high uncertainty in the initial conditions. The spread in 
+    probability decreases as we add more measurement points. Readers would like
+    to draw similarities to Kalman filter, only in this case we can apply it to 
+    any non-linear continuous dynamical system.}
+    \label{fig:cnf}
 \end{figure}
@@
 \subsection{Robustness analysis of NODEs}
-
-\subsection{Efficient training techniques for NODEs}
 

### 2025-03-20 ABC_plot.py
forgot to add this

@@ -67,3 +67,18 @@ with PdfPages('ABC_plot.pdf') as pdf:
     pdf.savefig()
     plt.show()
     plt.close()
+
+    par = 10
+    plt.figure(figsize=figsize)
+    t = np.linspace(0, 1, 100)
+    exppos = np.exp(par*t)
+    expneg = np.exp(-par*t)
+    t = np.linspace(1, 0, 100)
+    plt.semilogy(t, exppos, '-', color='red', label='Exploding grad')
+    plt.semilogy(t, expneg, '-', color='blue', label='Vanishing grad')
+    plt.xlabel('$t$')
+    plt.ylabel('$\lambda$', rotation=0)
+    plt.legend()
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()

### 2025-03-20 paper.tex
completed the report finally

@@
 systems. While such systems can be approximated using finite impulse response (FIR) 
-or autoregressive exogenous (ARX) models, these approaches often fail to respect 
+or autoregressive exogenous (ARX) models \citep{ljung:1999}, these approaches often fail to respect 
 fundamental physical laws. One way to address this limitation is by penalizing 
@@
 rather as a solution strategy for PDEs and ODEs. Fundamentally, they operate 
-as \textit{methods of weighted residuals}—specifically, collocation schemes—where 
+as \textit{methods of weighted residuals} \citep{villadsen:stewart:1967}—specifically, collocation schemes—where 
 the residual is minimized at selected collocation points. While PINNs can be 
@@
 \end{align}
-To propagate the gradient we need to solve the sensitivity equations:
+We refer \cite{rawlings:ekerdt:2020} for following derivation. To propagate the gradient we need to solve the sensitivity equations:
 \begin{align}
@@
 In this section we redefine the objective function as an integral over time.
-The integral formulation allows us to derive the adjoint equation.
+The integral formulation allows us to derive the adjoint equation. We refer \cite{sengupta:friston:penny:2014} 
+that gives a straightforward proof of the adjoint equation than the one in
+\cite{chen:rubanova:bettencourt:duvenaud:2018}.
 Consider following least-squares objective for \cref{eq:ssode} and the Lagrange functional:
@@
 \dfrac{\partial \mathcal{L}}{\partial \theta^T}$.
-The original NODE paper uses objective function of the form:
+\cite{chen:rubanova:bettencourt:duvenaud:2018} uses objective function of the form:
 \begin{align}
@@
 We can evaluate the gradient in one go as shown in \cref{eq:adjointnode}. 
-Some recent solutions to storing intermediate points involve \textit{checkpointing}
-at only few points. However, such a method is not 
+However, such method is not 
 an ultimate panacea. Such systems are still vulnerable to the problems commonly faced by
@@
 at different time points. Sometimes we may end up with noisy gradients which may lead to poor convergence.
+Some recent remedies is solving adjoint equation in multiple time-intervals rather than 
+a single long time-interval
+and storing intermediate points, often called as \textit{checkpointing}.
+ \citep{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}.
 
@@
 
-The NODE paper applies the same motivation to develop continuous normalizing flows.
-Interested reader can refer the paper for theorem on instantaneous change of variables.
+The NODE paper applies the same motivation to develop continuous normalizing flows
+\citep{grathwohl:chen:bettencourt:sutskever:duvenaud:2018}.
+Interested reader can refer the paper for proof of theorem on instantaneous change of variables
+\citep{chen:rubanova:bettencourt:duvenaud:2018}.
 The key idea is the neural ODE rhs is trained to conserve the probability density 
@@
 \end{align}
-The author would like to draw parallels to the classical Kalman filter that also does the similar job,
+The author would like to draw parallels to the classical Kalman filter \citep{kalman:1960} that also does the similar job,
 but the continuous normalizing flow can be applied to any non-linear dynamical system.
@@
 
+Well-posedness of an IVP is given by following theorem \citep{ascher:petzold:1998}:
+\begin{theorem}
+    Let $f(t,x)$ be continuous for all $(t,x)$ in a region $\mathcal{D} = {0 \leq t \leq b, |x| \leq \infty}$. 
+    Moreover assume Lipschitz contnuity in $x$, then there exists a constant $L$ such that:
+    \begin{align*}
+        \norm{f(t,x) - f(t,\bar{x})} \leq L \norm{x-\bar{x}}
+    \end{align*}   
+\end{theorem}
+Thus, for ODEs with slightly perturbed initial conditions, the solution should not
+intersect though they can approach the same trajectory. Thus, the solution of an 
+ODE can be sandwhiched between two other perturbed solutions. Refer to \cref{fig:robust} for
+a simple example that again follows the kinetics in \cref{eq:atobtoc}.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.55\textwidth, page=5]{ABC_plot.pdf} 
+    \caption{Robustness analysis of ODEs. We see that the solution $c_{B2}$ 
+    is sandwiched between two perturbed solutions $c_{B1}, c_{B3}$, 
+    though all approach the same steady-state. Also note that the slightly time-shifted
+    solution appears as a perturbed solution}
+    \label{fig:robust}
+\end{figure}
+Thus, following above idea, \cite{yan:du:tan:feng:2019} developed robust NODEs.
+The idea is the perturbed solution can be seen as time-shifted solution of the
+original ODE sytem. Using uniqueness of solutions, time-invariant NODEs can be regularized by
+adding a penalty term to the loss function. The penalty term is given by:
+\begin{align*}
+    \norm{x(t+T)-x(t)} \leq \norm{\int_t^{t+T} f_{NN}(x, u; \theta) dt}
+\end{align*}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.55\textwidth, page=6]{ABC_plot.pdf} 
+    \caption{If we train based only on prediction loss, the model can overfit.
+    However, if we penalize the jacobian of the loss, we can regularize the model
+    to damp the oscillations.}
+    \label{fig:jacobnode}
+\end{figure}
+
+\cite{finlay:jacobsen:nurbekyan:oberman:2020} recommend to penalize the Jacobian of ODE rhs function
+$(\nabla_x f)$ to damp oscillations that happen due to overfitting in neural ODE. A simple example
+using polynomial fit is show in \cref{fig:jacobnode}.
 \section{Conclusions}
+In this review, we explored the mathematical foundations and practical applications 
+of Neural Ordinary Differential Equations (NODEs), highlighting their connection 
+to classical ODE-based modeling and modern machine learning approaches. We first 
+established how NODEs extend traditional discrete architectures such as ResNets 
+and RNNs by formulating neural networks as continuous dynamical systems. This 
+formulation provides a more principled way to model physical and process-driven 
+data while reducing memory constraints through adjoint-based sensitivity analysis.
+
+A key focus of this review was the gradient computation techniques for NODEs. We 
+discussed the forward sensitivity method, which requires solving an augmented ODE 
+system but suffers from an increase in computational cost with the number of parameters. 
+The adjoint sensitivity method, in contrast, offers a more efficient alternative by 
+solving a backward ODE, making NODEs feasible for large-scale learning problems. While 
+adjoint methods significantly improve efficiency, they are not free from challenges, 
+as stiffness and numerical instabilities can still affect convergence.
+
+We also explored continuous normalizing flows (CNFs), where NODEs are utilized to 
+construct generative models that explicitly model probability density evolution. 
+The application of CNFs to chemical kinetics demonstrated how this framework can 
+be extended to physical modeling while drawing conceptual parallels with classical 
+filtering techniques such as the Kalman filter. These methods provide powerful tools 
+for uncertainty quantification and data-driven discovery in dynamical systems.
+
+A critical aspect of NODE-based learning is the robustness of solutions. We examined 
+how NODEs, as solutions to differential equations, inherit the stability properties 
+of the underlying ODEs. In particular, well-posedness ensures that small perturbations 
+in initial conditions do not lead to solution intersections, reinforcing the interpretability 
+and reliability of NODE-based models. This insight was leveraged to propose robust NODEs, 
+where regularization techniques enforce time-invariance by penalizing deviations between time-shifted trajectories.
+
+Overall, NODEs present a compelling framework for modeling dynamical systems in a continuous, 
+memory-efficient, and theoretically grounded manner.
+
+\section{Data availability and Code}
+All the figures in the review have been generated by the author. The code for the
+figures can be found at
 

### 2025-03-20 ABC_plot.py

@@ -5,6 +5,8 @@ import matplotlib.pyplot as plt
 import numpy as np
 from matplotlib.backends.backend_pdf import PdfPages
 import pickle
+from scipy.linalg import expm
+from scipy.linalg import lstsq
 
 figsize = (4,3)
 data = pickle.load(open('ABC_solve.pickle', 'rb'))
@@ -65,7 +67,6 @@ with PdfPages('ABC_plot.pdf') as pdf:
 
     fig.subplots_adjust(left=0.05, right=0.99, bottom=0.05, top=0.99, wspace=0.05, hspace=0.1)
     pdf.savefig()
-    plt.show()
     plt.close()
 
     par = 10
@@ -78,6 +79,87 @@ with PdfPages('ABC_plot.pdf') as pdf:
     plt.semilogy(t, expneg, '-', color='blue', label='Vanishing grad')
     plt.xlabel('$t$')
     plt.ylabel('$\lambda$', rotation=0)
+    plt.legend()
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()
+
+    # K = np.array([[-1, 1/3], [1, -(1/3+0.1)]])
+    # t = np.linspace(0, 50, 200)
+    # eps = 0.1
+    # c0 = np.array([1 + eps, 0 + eps])
+    # c0_perturb = np.array([1 + 2*eps, 0 + 2*eps])
+    # c0_perturb2 = np.array([1 + 3*eps, 0 + 3*eps])
+    # c = np.array([c0 @ expm(K*t) for t in t])
+    # c_perturb = np.array([c0_perturb @ expm(K*t) for t in t])
+    # c_perturb2 = np.array([c0_perturb2 @ expm(K*t) for t in t])
+
+    # plt.figure(figsize=figsize)
+    # plt.plot(t, c[:,1], '-', color='red', label='$c_{B1}$')
+    # plt.plot(t, c_perturb[:,1], '--', color='blue', label='$c_{B2}$')
+    # plt.plot(t, c_perturb2[:,1], color='green', label='$c_{B3}$')
+    # plt.plot(t[:-2], c[2:,1], color='black', label='$c_{B1}(t + T)$, T=0.5')
+
+    # plt.xlabel('$t$')
+    # plt.ylabel('$c_B$', rotation=0)
+    # plt.legend()
+    # plt.tight_layout()
+    # pdf.savefig()
+    # plt.close()
+
+    K = np.array([-1])
+    t = np.linspace(0, 2, 75)
+    eps = 0.1
[…]

### 2025-03-20 paper.tex
some adjustments

@@
 Department of Chemical Engineering, University of California, Santa Barbara}
-\date{\textit{ME255NN: Neural Networks for modeling, optimization and control} \\
+\date{\today\\
+\textit{ME255NN: Neural Networks for modeling, optimization and control} \\
 \textit{Winter 2025}}
@@
 developments in the field.
+
+\noindent \textbf{Notation}: We keep our notation consistent with the one used in
+\cite{ rawlings:mayne:diehl:2020}.
 
@@
         f_{NN}(x, u; \theta) \\
-        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
+        -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
     \end{bmatrix} dt;
@@
         k_{2}c_B\\
-        -\mathrm{Tr} \left(\dfrac{\partial f_{NN}}{\partial x^T}\right)
+        -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial c^T}\right)
     \end{bmatrix};
@@
     \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
-    \min\limits_{\theta} V(x) &= -\log p(x) + \norm{c - \tilde{c}}^2
+    \min\limits_{\theta} V(c) &= -\log p(c) + \norm{c - \tilde{c}}^2
 \end{align}
@@
 All the figures in the review have been generated by the author. The code for the
-figures can be found at
+figures can be found at \url{https://github.com/dakeprithvi/2025a_cnf}.
 

### 2025-03-20 talk.tex
make final touches to talk

\documentclass[dvipsnames, 9pt]{beamer}

%\documentclass[xcolor=dvipsnames, 8pt]{beamer} %
%\setbeamertemplate{navigation symbols}{}

\usetheme{SantaBarbara}

%light gray/black highlighting to be used with pause command
\colorlet{shadecolor}{gray!40}
\def\blackgray<#1>{%
  \temporal<#1>{\color{shadecolor}}{\color{black}}{\color{shadecolor}}}

\definecolor{black}{HTML}{0A0A0A}
\definecolor{red}{HTML}{e00404}

\definecolor{blue}{HTML}{0647A8}
\definecolor{darkgreen}{HTML}{008000}

\definecolor{Asparagus}{HTML}{87A96B}

\usepackage[utf8]{inputenc}
\usepackage{algorithm, algpseudocode}

\usepackage{tightlist}
\usepackage{tikz, tikzsettings}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\pdfmapfile{+sansmathaccent.map} % Fix done for making the talk in ubuntu.
\graphicspath{{./}{./figures/}{./figures/presentation/}}
\usepackage{makecell}
\usepackage{booktabs}

\usepackage{subcaption}
\usepackage[authoryear,round]{natbib}
\usepackage{color}
\usepackage{colortbl}
\usepackage{xcolor}
\usepackage{pgfplots}
\usepackage{ragged2e}
\usepackage{rxn}
\usepackage{hyperref}

\usetikzlibrary{shapes,calc,spy, calc, backgrounds,arrows, fit, decorations.pathmorphing, decorations.pathreplacing, matrix}
\usepackage{caption}
\usepackage{mpcsymbols}
\usepackage{graphicx}

\newcommand{\calert}[1]{\textcolor{blue}{#1}}

\makeatother
\setbeamertemplate{footline}
{\leavevmode%
	\hbox{%
		\begin{beamercolorbox}[wd=.3\paperwidth,ht=2.25ex,dp=1ex,center]{author
		in head/foot}%
			\usebeamerfont{author in head/foot}\insertshortauthor
		\end{beamercolorbox}%
		\begin{beamercolorbox}[wd=.6\paperwidth,ht=2.25ex,dp=1ex,center]{title
		in head/foot}%
			\usebeamerfont{title in head/foot}\insertshorttitle
		\end{beamercolorbox}%
		\begin{beamercolorbox}[wd=.1\paperwidth,ht=2.25ex,dp=1ex,center]{date in
		head/foot}%
			\insertframenumber{} / \inserttotalframenumber\hspace*{1ex}
	\end{beamercolorbox}}%
	\vskip0pt%
}

\renewcommand{\vec}{\textnormal{vec}} 
\newcommand{\stoi}{\text{\boldmath $\nu$}}

\title[NODEs]{Stirring Up Neural ODEs: A Chemical Engineer's Recipe for Dynamic Modeling
}

\author[ME255NN--Dake]{Prithvi Dake}
\institute [UCSB]{Department of Chemical
Engineering\\
\pgfuseimage{ucsb-logo}}

\date{ME255NN Neural ordinary differential equations  \\
\today}

%\AtBeginSection[]
%{
%  \begin{frame}
%    \frametitle{Outline}
%    \tableofcontents[currentsection]
%  \end{frame}}

\begin{document}

\frame{\titlepage}

\begin{frame}{Gentle intro with a simple problem}
    \begin{columns}
    {\begin{column}{0.45\textwidth}
    \begin{block}{}
        Consider following \textcolor{red}{\textit{true}} model of linear chemical kinetics taking place
        in a well-mixed batch reactor with $c = (c_A, c_B, c_C)^T$ and initial concentration
        $c_0$:
        \begin{rxn*}{} 
        A \rlh[k_1][k_{-1}] 2 B, \qquad  B \rarrow[k_2] C
        \label{rxn:atobtoc}
        \end{rxn*}
        \begin{gather*}
        \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \end{bmatrix}
        =\begin{bmatrix}
            -k_{1} & k_{-1} & 0 \\
            2k_{1} & 2k_{-1} - k_2 & 0 \\
            0 & k_{2} & 0
        \end{bmatrix}
        \begin{bmatrix}  c_A \\ c_B \\ c_C \end{bmatrix} 
        \label{eq:atobtoc}
        \end{gather*}
    \end{block}
    \end{column}}
    {\begin{column}{0.5\textwidth}
    \begin{figure}[h]
        \centering
        \includegraphics[width=\textwidth, page=7]{ABC_plot.pdf} 
    \end{figure}
    \end{column}}
    \end{columns}
\end{frame}

\begin{frame}{Background: If PINNs why NODEs? }
    \textbf{Training a Physics-Informed Neural Network (PINN) \footnote{\cite{raissi:perdikaris:karniadakis:2019}}}
    \begin{algorithmic}[1]
        \State \textbf{Input:} Neural network architecture $f_{NN}$
        \State \textbf{Physical Model:} $Lx=f \quad B_1 x = 0 \quad B_2 x = 0$
        \Comment{True or known physics}
        \State \textbf{Initialize} neural network parameters $\theta$
        \Repeat
            \State $x = f_{NN}(x,t; \theta)$
            \State $V_{ODE} = \norm{Lx - f}^{2}$ 
            \Comment{Prediction error minimization} 
            \State $V_{BC} = \norm{B_1 x}^{2} + \norm{B_2 x}^{2}$
            \Comment{Enforce boundary conditions}
            \State $V = \lambda_1 V_{PDE}  + \lambda_2 V_{BC}$
            \Comment{Combine loss terms}
            \State $\theta \leftarrow \theta - \eta \nabla_{\theta} V$
            \Comment{Update $\theta$ using gradient descent}
        \Until{convergence criteria is met}
        \State \textbf{Return} trained model $f_{NN}$
    \end{algorithmic}
\end{frame}

\begin{frame}{Background: If PINNs why NODEs?}
    \begin{columns}
    {\begin{column}{0.5\textwidth}
    \begin{figure}[h]
        \centering
        \includegraphics[width=\textwidth, page=8]{ABC_plot.pdf} 
    \end{figure}
    \end{column}}
    {\begin{column}{0.45\textwidth}
    \begin{block}{}
        \begin{itemize}  % Right triangle symbol
            \item I view PINNs as a softer version\footnote[frame]{By softer I 
            mean PINNs minimize rather than exactly satisfy equality constraints } of \textit{method of weighted residuals}.
            \footnote[frame]{\cite{villadsen:stewart:1967}}
            \item Mostly used as surrogate numerical solver for PDEs.
            \item Can learn some physics but has no \textit{structure}.
            \item Causality problem (no sequential model, future depends on past).
            \footnote[frame]{\cite{wang:shyam:paris:2022}}
        \end{itemize}
    \end{block}
    \begin{block}{Recent advances}
        Physics-informed neural ODE (PINODE) \footnote[frame]{\cite{sholokhov:liu:mansour:nabi:2023}} \\
        Lift-and-embed: Koopman operators \footnote[frame]{\cite{liu:sholokov:mansour:nabi:2024}}
    \end{block}
    \end{column}}
    \end{columns}
\end{frame}

\begin{frame}{ResNets and RNNs}
    \begin{block}{A state-space representation: Euler discretized system}
    \begin{align*}
        x^+ &= x + f_{NN} (x, u; \theta) &\qquad x^+ = f_{NN} (x, u; \theta)\notag \\ 
        y &= g_{NN} (x, u; \theta) &\qquad y = g_{NN} (x, u; \theta)
        \label{eq:resrnn}
    \end{align*}
    Parallels with FIR and ARX models in system identification \footnote[frame]{\cite{ljung:1999}}.\\
    Backprop strategy: Brute-force chain rule (store intermediate states).
    \end{block}
    \begin{columns}
    {\begin{column}{0.6\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth, page=2]{ABC_plot.pdf}
    \end{figure}
\end{column}}
{\begin{column}{0.3\textwidth}
    \ If $x\in\R^n, \theta\in R^p$, $S = \dfrac{\partial x}{\partial \theta^T}\in\R^{n\times \textcolor{red}{p}}$
\end{column}}
\end{columns}
\end{frame}

\begin{frame}{Don't want to store intermediate states? Forward sensitivities for IVPs \footnote{Derivation in \cite{rawlings:ekerdt:2020}}}
    \begin{itemize}
    \item Consider a non-linear state-space model ($y=x$):
    \begin{align*}
        \dot{x} &= f(x, u; \theta) \qquad x(0) = g(x_0; \theta)
    \end{align*}
    \item Sensitivity is defined as and evolves as \footnote{Using Clairaut's Theorem on Mixed Partials}:
    \begin{align*}
        S &= \dfrac{\partial x}{\partial \theta^T} \\
        \dfrac{d S}{dt} &= \dfrac{d}{dt} \left(\dfrac{\partial x}{\partial \theta^T}\right) =
        \dfrac{\partial}{\partial\theta^T} \left(\dfrac{d x}{dt}\right) = 
        \dfrac{\partial }{\partial \theta^T} f
    \end{align*}
    \item Augmented system of ODEs:
    \begin{align*}
        \begin{bmatrix} \dot{x} \\ \dot{S} \end{bmatrix} = 
        \begin{bmatrix} f \\ \dfrac{\partial f}{\partial x^T}S + \dfrac{\partial f}{\partial \theta^T} \end{bmatrix}
        \qquad
        \begin{bmatrix} x(0) \\ S(0) \end{bmatrix} =
        \begin{bmatrix} x_0 \\ \dfrac{\partial g}{\partial \theta^T} \end{bmatrix}
    \end{align*}
    \item Notice again: $S\in \R^{n\times p}$ i.e sensitivities we need to solve
    increase \textcolor{red}{linearly} with $p$. Imagine how bad this could get for FNN with millions of parameters and 
    states!
    \end{itemize}
\end{frame}

\begin{frame}{How to avoid the linear cost? Pontryagin's adjoint method}
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{oldies.png}\\
        \textit{One mastered control, one shaped space, one gambled with certainty}.\\
    \end{figure}
\end{frame}

\begin{frame}{Backward adjoint method}
    \begin{itemize}
        \item Consider following optimization problem and the Lagrange functional:
        \begin{align*}
            \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \hat{x}}^2 dt = \int_0^t g(x)dt \notag \\
            \min\limits_{\theta, \lambda}\mathcal{L} &= \int_0^t g dt + \int_0^t \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
        \end{align*}
        \item The augmented system of ODEs is \footnote{Refer \cite{sengupta:friston:penny:2014} for paper and 
        \url{https://youtu.be/k6s2G5MZv-I?si=ZIyP7MvgzTK7D9kn} for a walkthrough. Additionally, 
        I have worked out the derivation step-by-step in my report.}:
        \[
        \begin{bmatrix} \dot{x} \\ \dot{\lambda^T} \end{bmatrix} = 
        \begin{bmatrix} f_{NN} \\ -\lambda^T \dfrac{\partial f_{NN}}{\partial x^T}\end{bmatrix}
        \qquad
        \begin{bmatrix} x(t) \\ \lambda^T(t) \end{bmatrix} =
        \begin{bmatrix} x_t \\ - \dfrac{\partial g}{\partial x^T}|_{t} \end{bmatrix}
        \label{eq:augsens}
        \]
        \item The gradient of loss wrt parameters is calculated as:
        \[
            \dfrac{\partial V}{\partial \theta^T} = \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt
        \]
    \end{itemize}
\end{frame}

\begin{frame}{An ultimate panacea?}
    \begin{columns}
    {\begin{column}{0.45\textwidth}
    \begin{figure}
        \centering
        \includegraphics[width=\textwidth, page=4]{ABC_plot.pdf}
    \end{figure}
    \end{column}}
    {\begin{column}{0.5\textwidth}
    \begin{block}{}
        \begin{itemize}
            \item Consider $\dot{\lambda} = \pm 10 \lambda$ and $\lambda(0) = 0$
            \item We see that the adjoint variable will either explode or vanish.
            This heavily impacts the learning problem. 
            \item The Jacobian $\dfrac{\partial f}{\partial x^T}$ can be ill-conditioned.
            \item We still solve the adjoint equation by quadrature, thus numerical noise 
            can still be a problem.
        \end{itemize}
    \end{block}
    \begin{block}{Recent remedies}
        \begin{itemize}
            \item Don't solve adjoint over long horizons, but rather 
            small pieces and store the initial conditions for those intermediate
            pieces.
            \item Popularly called as \textit{checkpointing} \footnote[frame]{\cite{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}}.
        \end{itemize}
    \end{block}
    \end{column}}
    \end{columns}
\end{frame}

\begin{frame}{NODE}
    \begin{algorithmic}[1]
        \State \textbf{Input:} Neural network architecture $f_{NN}$
        \State \textbf{ODE:} 
        $
            \begin{bmatrix} \dot{x} \\ \dot{\lambda^T} \end{bmatrix} = 
            \begin{bmatrix} f_{NN} \\ -\lambda^T \dfrac{\partial f_{NN}}{\partial x^T}\end{bmatrix}
            \qquad
            \begin{bmatrix} x(t) \\ \lambda^T(t) \end{bmatrix} =
            \begin{bmatrix} x_t \\ - \dfrac{\partial g}{\partial x^T}|_{t} \end{bmatrix}
            \label{eq:augsens}
        $
        \Comment{Construct the augmented ODE using \texttt{PyTorch}}
        \State \textbf{Initialize} neural network parameters $\theta$
        \Repeat
            \State $V = \norm{\hat{x} - x}^2$
            \Comment{Forward solve to get the loss $V$}
            \State $\hat{x}_{aug} = \text{ODESolve}(f_{aug}, x_{aug}(0), T, t_0)$
            \State $\dfrac{\partial V}{\partial \theta^T} = \lambda^T(0)\dfrac{\partial x}{\partial \theta^T}|_0 + \int_0^t   \lambda^T \dfrac{\partial f}{\partial \theta^T} dt$
            \Comment{Construct the gradient wrt loss}
            \State $\theta \leftarrow \theta - \eta \nabla_{\theta} V$
        \Until{convergence criteria is met}
        \State \textbf{Return} trained model $f_{NN}$
        \end{algorithmic}        
\end{frame}

\begin{frame}{NODE}
    \begin{figure}
        \centering
        \includegraphics[width=0.7\textwidth, page=1]{ABC_plot.pdf}\\
        Concentration profiles of species A, B, and C in a batch reactor. The
    NODE model also does quite a good job to predict the concentration profiles. 
    In fact for such simple problem, we hardly see any difference between ResNet and NODE.
    \end{figure}
\end{frame}

\begin{frame}{Generative latent model using CNF (Generative=Probabilistic)}
\end{frame}

\begin{frame}[allowframebreaks]{References}
	\renewcommand{\refname}{}
	\bibliographystyle{abbrvnat}
	%\bibliographystyle{abbrvnat_no_url}
	\bibliography{abbreviations,articles,books,proceedings, cnf_bib}
\end{frame}

\end{document}

### 2025-03-20 ABC_plot.py

@@ -14,6 +14,7 @@ t = data['t']
 y_meas = data['y_meas']
 y_pred = data['y_pred']
 y_pred_res = data['y_pred_res']
+y_pred_pnn = data['y_pnn']
 y = data['y']
 y[:,:,3] = np.exp(y[:,:,3])
 
@@ -164,3 +165,30 @@ with PdfPages('ABC_plot.pdf') as pdf:
     plt.tight_layout()
     pdf.savefig()
     plt.close()
+
+    plt.figure(figsize=figsize)
+    t = data['t']
+    plt.plot(t, y_meas[:,0], 'o', color='red', mfc='none', label='$c_A$')
+    plt.plot(t, y_meas[:,1], 'o', color='blue', mfc='none', label='$c_B$')
+    plt.plot(t, y_meas[:,2], 'o', color='green', mfc='none', label='$c_C$')
+    plt.xlabel('$t$')
+    plt.ylabel('$C$', rotation=0)
+    plt.legend(loc='lower right')
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()
+
+    plt.figure(figsize=figsize)
+    plt.plot(t, y_meas[:,0], 'o', color='red', mfc='none')
+    plt.plot(t, y_pred_pnn[:,0], '-', color='red')
+    plt.plot(t, y_meas[:,1], 'o', color='blue', mfc='none')
+    plt.plot(t, y_pred_pnn[:,1], '-', color='blue')
+    plt.plot(t, y_meas[:,2], 'o', color='green', mfc='none')
+    plt.plot(t, y_pred_pnn[:,2], '-', color='green')
+    plt.xlabel('$t$')
+    plt.ylabel('$C$', rotation=0)
+    plt.tight_layout()
+    pdf.savefig()
+    plt.close()
+
+    

### 2025-03-20 ABC_solve.py

@@ -23,12 +23,16 @@ abc.multi_solve(ntimes=10)
 # train the CNF
 abc.cnf_train(niter=500)
 
+
+abc.pnn_train()
+
 data = {}
 data['t'] = t
 data['y_meas'] = y_meas
 data['y_pred'] = y_pred
 data['y_pred_res'] = y_pred_res
 data['y'] = abc.y
+data['y_pnn'] = abc.c_pred
 
 with open('ABC_solve.pickle', 'wb') as f:
     pickle.dump(data, f)

### 2025-03-20 lib/ABC_plant.py

@@ -72,6 +72,13 @@ class ABC(nn.Module):
         )
         self.cnf.apply(self.init_weights)
 
+        self.pnn = nn.Sequential(
+            nn.Linear(1, 8),
+            nn.Tanh(),
+            nn.Linear(8, 3)
+        )
+        self.pnn.apply(self.init_weights)
+
      
     @staticmethod
     def init_weights(m):
@@ -311,6 +318,47 @@ class ABC(nn.Module):
         toc = time.time()
         print(f'Training time: {(toc - tic)/60} mins')
         self.y = self.y.detach().numpy()
+
+    
+    def pnn_loss(self):
+        '''
+        Compute the PINN loss for fitting the kinetics ODE.
+        '''
+        t = torch.linspace(self.t_start, self.t_end, self.nplot, dtype=self.dtype, requires_grad=True)
+
+        c_pred = self.pnn(t.reshape(-1,1))
+        c_a, c_b, c_c = c_pred[:, 0], c_pred[:, 1], c_pred[:, 2]
+
+        dca_dt = torch.autograd.grad(c_a, t, grad_outputs=torch.ones_like(c_a), create_graph=True)[0]
+        dcb_dt = torch.autograd.grad(c_b, t, grad_outputs=torch.ones_like(c_b), create_graph=True)[0]
+        dcc_dt = torch.autograd.grad(c_c, t, grad_outputs=torch.ones_like(c_c), create_graph=True)[0]
+
+        ode1 = torch.mean((dca_dt - (-self.k1 * c_a + self.k_1 * c_b))**2)
+        ode2 = torch.mean((dcb_dt - (2 * self.k1 * c_a - (2 * self.k_1 + self.k2) * c_b))**2)  
+        ode3 = torch.mean((dcc_dt - (self.k2 * c_b))**2)
+
+        pred_loss = torch.mean((c_pred - self.measure)**2)
+        self.loss_for_pnn = ode1 + ode2 + ode3 + pred_loss
+        self.c_pred = c_pred
+        return self.loss_for_pnn
+    
+    def pnn_train(self, niter = 1000):
+        '''
+        Train the PINN
+        '''
+        optimizer = optim.Adam(self.pnn.parameters(), lr=1e-2)
+        tic = time.time()
+        for i in range(niter):
+            optimizer.zero_grad()
+            loss = self.pnn_loss()
+            loss.backward()
+            optimizer.step()
+            if i % 100 == 0:
[…]

### 2025-03-20 talk.tex
finally done with the talk

@@
 
-\title[NODEs]{Stirring Up Neural ODEs: A Chemical Engineer's Recipe for Dynamic Modeling
-}
+\title[NODEs]{Stirring Up Neural ODEs: A Chemical Engineer's perspective}
 
@@
 
-\date{ME255NN Neural ordinary differential equations  \\
+\date{ME255NN Report  \\
 \today}
@@
 {\begin{column}{0.3\textwidth}
-    \ If $x\in\R^n, \theta\in R^p$, $S = \dfrac{\partial x}{\partial \theta^T}\in\R^{n\times \textcolor{red}{p}}$
+    \ If $x\in\R^n, \theta\in \R^p$, $S = \dfrac{\partial x}{\partial \theta^T}\in\R^{n\times \textcolor{red}{p}}$
 \end{column}}
@@
     \item Notice again: $S\in \R^{n\times p}$ i.e sensitivities we need to solve
-    increase \textcolor{red}{linearly} with $p$. Imagine how bad this could get for FNN with millions of parameters and 
+    for increase \textcolor{red}{linearly} with $p$. Imagine how bad this could get for FNN with millions of parameters and 
     states!
@@
         \includegraphics[width=0.8\textwidth]{oldies.png}\\
-        \textit{One mastered control, one shaped space, one gambled with certainty}.\\
+        %\textit{One mastered control, one shaped space, one gambled with certainty}.\\
     \end{figure}
@@
     \begin{itemize}
-        \item Consider following optimization problem and the Lagrange functional:
+        \item Consider following optimization problem and the Lagrange functional given data $\tilde{x}$:
         \begin{align*}
-            \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \hat{x}}^2 dt = \int_0^t g(x)dt \notag \\
+            \min\limits_{\theta}V(x) &= \int_0^t \norm{x - \tilde{x}}^2 dt = \int_0^t g(x)dt \notag \\
             \min\limits_{\theta, \lambda}\mathcal{L} &= \int_0^t g dt + \int_0^t \lambda(t)^T \left(f - \dfrac{dx}{dt}\right) dt
@@
             \item Consider $\dot{\lambda} = \pm 10 \lambda$ and $\lambda(0) = 0$
-            \item We see that the adjoint variable will either explode or vanish.
+            \item We see that the adjoint variable will either \textit{explode} or 
+            \textit{vanish}.
             This heavily impacts the learning problem. 
             \item The Jacobian $\dfrac{\partial f}{\partial x^T}$ can be ill-conditioned.
-            \item We still solve the adjoint equation by quadrature, thus numerical noise 
+            \item We solve the adjoint equation by quadrature, thus numerical noise 
             can still be a problem.
@@
     \end{block}
-    \begin{block}{Recent remedies}
+    \begin{block}{Remedies}
         \begin{itemize}
             \item Don't solve adjoint over long horizons, but rather 
-            small pieces and store the initial conditions for those intermediate
-            pieces.
+            small windows and store the initial conditions for those intermediate
+            windows.
             \item Popularly called as \textit{checkpointing} \footnote[frame]{\cite{zhuang:dvornek:li:tatikonda:papademetris:duncan:2020}}.
@@
         \centering
-        \includegraphics[width=0.7\textwidth, page=1]{ABC_plot.pdf}\\
+        \includegraphics[width=0.65\textwidth, page=1]{ABC_plot.pdf}\\
         Concentration profiles of species A, B, and C in a batch reactor. The
     NODE model also does quite a good job to predict the concentration profiles. 
-    In fact for such simple problem, we hardly see any difference between ResNet and NODE.
+    In fact for such simple problem, we hardly see any difference between ResNet and NODE
+    \footnote{All the case studies in this report are built using \texttt{Torchdiffeq} by \cite{torchdiffeq}}.
     \end{figure}
@@
 
-\begin{frame}{Generative latent model using CNF (Generative=Probabilistic)}
+\begin{frame}{Generative latent model using CNF}
+    \begin{itemize}
+        \item \textbf{Latent} = States in state-space model, since we usually cannot measure
+        them directly. Thus, we construct state-estimators.
+        \item \textbf{Generative latent} = State-space model that also propagates uncertainty
+        in estimates i.e. you can sample and \textit{generate} multiple trajectories after learning.
+        \item Surprise! Good old Kalman filter can be called a generative latent model!
+        \footnote{Refer \cite{rawlings:mayne:diehl:2020} for derivation of the Kalman filter using
+        Bayesian estimation principles.}
+        \item Consider a simple linear transform:
+        \begin{align*}
+            x &\in \mathcal{N}(m, \sigma^2 I) \\
+            y &= Ax \\
+            y &\in \mathcal{N}(Am, A\sigma^2 A^T)
+        \end{align*}
+        \item Bayesian approach: $\textcolor{blue}{p(y|x)} \sim p(x|y) \textcolor{red}{p(y)}$.
+        Assuming an \textcolor{red}{uniform prior} the likelihood problem is 
+        \[
+        \theta = \arg\max\limits_{\theta} \log p(x|y; \theta) \qquad p(x|y) = A^{-1} y \sim \mathcal{N}(m, \sigma^2I)        \]
+        \item Ofcourse, $A$ is assumed to be non-singular (i.e. unique mapping between $x$ and $y$;
+        \textit{bijective mapping}).
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Extending the bayesian framework for any non-linear transform}
+    \begin{itemize}
+        \item Consider following series of transformations (\textit{planar-flow}):
+        \[
+        x^+ = f_{NN}(x, u; \theta) \qquad x(0) \sim \mathcal{N}(m, \sigma^2I) \qquad x(t) \sim ?
+        \]
+        \item Use \textit{change of variables}, compute the
+        \textcolor{red}{determinant} of Jacobian ($\sim N(o^3)$) and conserve the probability mass as well
+        as maximize the likelihood.
+        \item Instead use continuous framework (\textit{continuous normalizing flow}) \footnote{Refer \cite{chen:rubanova:bettencourt:duvenaud:2018} for proof}:
+        \begin{align*}
+        \dfrac{dx}{dt} = f_{NN}(x, u; \theta) &\qquad x(0) \sim \mathcal{N}(m, \sigma^2I) \qquad x(t) \sim ?\\
+        \dfrac{\partial p(x)}{\partial t} &=  -\mathrm{tr}\nabla_x f_{NN}
+        \end{align*}
+        \item Cost to get the trace of Jacobian is now linear $\sim N(o)$
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Generative latent model for kinetics}
+    \begin{figure}
+        \centering
+        \includegraphics[width=\textwidth, page=3]{ABC_plot.pdf}\\
+    \end{figure}
+    \begin{align*}
+        \begin{bmatrix} \dot{c}_A \\ \dot{c}_B \\ \dot{c}_C \\ \dot{\log p} \end{bmatrix}
+        = \begin{bmatrix}
+            -k_{1}c_A + k_{-1}c_B  \\
+            2k_{1}c_A +  2k_{-1}c_B - k_2c_B \\
+            k_{2}c_B\\
+            -\mathrm{tr} \left(\dfrac{\partial f_{NN}}{\partial c^T}\right)
+        \end{bmatrix};
+        &\qquad
+        \begin{bmatrix} c_A(0) \\ c_B(0) \\ c_C(0) \\ \log p(0) \end{bmatrix} =
+        \begin{bmatrix} 1 \\ 0 \\ 0 \\ \log{\mathcal{N}([c_A, c_B, c_C]^T, \sigma^2I)} \end{bmatrix} \notag \\
+        \min\limits_{\theta} V(c) &= -\log p(c) + \norm{c - \tilde{c}}^2
+    \end{align*}
+\end{frame}
+
+\begin{frame}{Robustness analysis of NODEs \footnote{Refer \cite{yan:du:tan:feng:2019}}}
+    \begin{figure}[h]
+        \centering
+        \includegraphics[width=0.5\textwidth, page=5]{ABC_plot.pdf} \\
+        Robustness analysis of ODEs. We see that the solution $c_{B2}$ 
+        is sandwiched between two perturbed solutions $c_{B1}, c_{B3}$, 
+        though all approach the same steady-state. Also note that the slightly time-shifted
+        solution appears as a perturbed solution \footnote{Refer \cite{ascher:petzold:1998}
+        for well-posedness theorem of IVPs.}
+        \label{fig:robust}
+    \end{figure}
+    \center \textbf{Theoretical bound}
+    \begin{align*}
+        \norm{x(t+T)-x(t)} \leq \norm{\int_t^{t+T} f_{NN}(x, u; \theta) dt}
+    \end{align*} 
+\end{frame}
+
+\begin{frame}{How to train your NODE?}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.55\textwidth, page=6]{ABC_plot.pdf}  \\
+    If we train based only on prediction loss, the model can overfit.
+    However, if we penalize the Jacobian term ($\dfrac{\partial f_{NN}}{\partial x^T}$), we can regularize the model
+    to damp the oscillations \footnote{\cite{finlay:jacobsen:nurbekyan:oberman:2020} \\ 
+    Access the code for all figures at \url{https://github.com/dakeprithvi/2025a_cnf}}
+    \end{figure}
 \end{frame}
